---
title: "PS5-ggplot"
author: "tiff"
date: "2023-02-17"
output: html_document
---

Part 1 - Load and Check Data:

```{r, message=FALSE}
library(tidyverse)

gapminder <- read_delim("data/gapminder.csv.bz2")
```

The data contains `r nrow(gapminder)` observations and has `r ncol(gapminder)`
variables.

Sample of the dataset use knitr:: kable
```{r, message=FALSE}
gapminder %>% 
  sample_n(3) 

```

The sample looks good, there are some missing values "NA", but the rest is organized,
clear, and the obervations makes sense with the variable


Part 2 - Descriptive Statistics:


```{r, message=FALSE}
gapminder %>%
  select(iso3, iso2, name) %>% 
  summarise( iso3 = length(unique(iso3)),
            iso2 = length(unique(iso2)),
            name = length(unique(name)))
```


Part A
 
```{r, message=FALSE}
gapminder %>%
  group_by(iso2) %>% 
  summarise(iso2code = n_distinct(name)) %>% 
  arrange(desc(iso2code))

```

Part B

```{r, message=FALSE}
gapminder %>%
  group_by(name) %>% 
  summarise(iso3code = n_distinct(iso3)) %>% 
  arrange(desc(iso3code))
```



Min and Max years in these data

```{r, message=FALSE}
gapminder %>% 
  filter(!is.na (time)) %>% 
  select(time) %>%
  summarise( minimum_year = min(time), maximum_year = max(time))
  
```


Part 3 - CO2 Emmissions


```{r, message=FALSE}
gapminder %>% 
  group_by(time) %>% 
  filter(is.na(co2), is.na(co2_PC)) %>% 
  count() %>% 
  arrange(desc(n))

```

(different variables, same numbers for the Co2 and Co2_PC)

```{r, message=FALSE}
total_CO2<- gapminder %>% 
  group_by(iso3, time) %>% 
  filter(co2 != "NA", name != "NA") %>% 
  summarise(avgCO2 = mean(co2)) %>% 
  filter(iso3 == "USA" | iso3 == "CHN" | iso3 == "IND" | iso3 == "VNM" | iso3 == "KOR")
  
 

total_CO2 %>% 
  ggplot(aes(time, avgCO2, col = factor(iso3))) +
  geom_line() +
  labs(x = "Time (yr)", y = "Average CO2 Emmission (kt)", col = "Countries",
       title = "Average CO2 Emission Over Year")
```



The graph suggests that out of the 5 countries chosen, China on average emits the highest amount
of Co2 after around 2005. China has the greatest positive trend throughout the years with a spike
in emissions after year 2000. The US started with the greatest amount of average co2 emission but 
has not had any major spikes, and is number 2 after 2000. While India, Korea, and Vietnam
emits the least amount of Co2 throughout the years in this group. 




```{r, message=FALSE}
percapita <- gapminder %>%
  group_by(iso3, time) %>% 
  filter(co2_PC != "NA", name != "NA") %>%
  summarise(avgcapita = mean(co2_PC)) %>% 
  filter(iso3 == "USA" | iso3 == "CHN" | iso3 == "IND" | iso3 == "VNM" | iso3 == "KOR")


percapita %>%
  ggplot(aes(time, avgcapita, col = factor(iso3))) +
  geom_line() +
  labs(x = "Time (yr)", y = "Average CO2 Emmission Per Capita (kt)", col = "Countries", 
       title = "Average CO2 Emission CP Over Year")

```



The graph suggests that the USA has the greatest average Co2 emission per-capita from 1960 to
2019, however there is a steady negative trend with the US releasing less CO2 per-capita in 2019 
than compared to previous years. With the other countries, there is a positive trend where 
they have been increasing CO2 emissions per-capital. 





```{r, message=FALSE}
gapminder %>%
  group_by(region, time) %>% 
  filter(time == 1960 | time ==2016) %>% 
  filter(co2_PC != "NA", region !="NA") %>%
  summarise(avgCO2_PC = mean(co2_PC))

```

Comment: what do you see?**

Region is split into continents, the variable time is in order of years from 1960 to 2019, and 
the average co2 emission per capita typically has a positive trend over the years. Except for 
the Americas where it is has a negative trend.
*****




```{r, message=FALSE}
gapminder %>%
  group_by(region, time) %>% 
  filter(time == 1960 | time == 2016) %>% 
  filter(co2_PC != "NA", region !="NA") %>%
  summarise(avgCO2_PC = unique(mean(co2_PC))) %>% 
  
  ggplot() +
  geom_col(aes(x = region, y = avgCO2_PC, fill = factor(time)), position = "dodge") +
  labs(title = "Average CO2 PC emission per Continent", 
           x = "Continent", y = "Average CO2 PC Emission (kt)", fill = "Year")

```


The bar chart suggests that in Africa, Asia, Europe, and Oceania the average CO2 emission per-capita 
were higher in 2016 than compared to 1960. However, in the Americas, average CO2 emission per-capita
was lower than it wa in 1960.






Year 2019 does not work, switch year to 2016, Checked with other TAs during OH

```{r, message=FALSE}
gapminder %>% 
  group_by(region) %>% 
  filter(time == 2016) %>% 
  filter(name != "NA", co2_PC != "NA", region != "NA") %>% 
  select(name, co2_PC) %>% 
  filter(rank(co2_PC) < 4 | rank(desc(co2_PC)) < 4) %>% 
  mutate(rank = co2_PC) %>% 
  arrange(region, co2_PC)


```



Part 4 - GDP Per Capita


```{r}
gapminder %>% 
  group_by(name) %>% 
  select(region, time, GDP_PC, lifeExpectancy, totalPopulation) %>% 
  filter(time == 1960) %>% 
  filter(GDP_PC != "NA", lifeExpectancy != "NA", name != "NA", totalPopulation != "NA") %>% 
  ggplot() +
  geom_point(aes(lifeExpectancy, GDP_PC, color = region, size = totalPopulation)) +
  labs(title = "1960 GDP PerCap. vs Life Expectancy", x = "Life Expectancy (Yr)", 
       y = "GDP Per Capita (Constant 2010 US$)", color ="Continent", size = "Country \n Size")

```



The scatter plot suggests that as life expectancy increases in years, so does the GDP Per Capita. 
Additionally, the plot depicts where continents are on the scale. For instance, we can see
that Africa is towards the left end with lower life expectancy and lower GDP CP while Europe
is on the right side of the plot with higher life expectancy and higher GDP CP. It also specifies
how large a country is by the size of the dot. Larger dot indicates a larger country and smaller
dot indicates a smaller country. 




























