---
title: "PS5-ggplot"
author: "tiff"
date: "2023-02-17"
output: html_document
---

**Part 1 - Load and Check Data:**

**Question 1**
```{r, message=FALSE}
library(tidyverse)

gapminder <- read_delim("data/gapminder.csv.bz2")
```

**Question 2**
```{r, message=FALSE}
nrow(gapminder)
ncol(gapminder)
```


The data contains `r nrow(gapminder)` observations and has `r ncol(gapminder)`
variables.


**Question3**


```{r, message=FALSE}
gapminder %>% 
  sample_n(5) %>% 
  knitr::kable()

```



_The sample looks good, there are some missing values "NA", but the data is organized,_
_clear, and the observations makes sense with the variable._


**Part 2 - Descriptive Statistics:**

**Question 1**

```{r, message=FALSE}
gapminder %>%
  select(iso3, iso2, name) %>% 
  summarise( iso3 = length(unique(iso3)),
            iso2 = length(unique(iso2)),
            name = length(unique(name)))
```


**Question 2 Part A**
 
```{r, message=FALSE}
gapminder %>%
  group_by(iso2) %>% 
  summarise(iso2code = n_distinct(name)) %>% 
  arrange(desc(iso2code))

```

**Question 2 Part B**

```{r, message=FALSE}
gapminder %>%
  group_by(name) %>% 
  summarise(iso3code = n_distinct(iso3)) %>% 
  arrange(desc(iso3code))
```



**Question 3**

```{r, message=FALSE}
gapminder %>% 
  filter(!is.na (time)) %>% 
  select(time) %>%
  summarise( minimum_year = min(time), maximum_year = max(time))
  
```


**Part 3 - CO2 Emissions:**

**Question 1**

```{r, message=FALSE}
gapminder %>% 
  group_by(time) %>% 
  filter(is.na(co2), is.na(co2_PC)) %>% 
  count() %>% 
  arrange(desc(n))

```


_(different variables, same numbers for the Co2 and Co2_PC)_
When looking at Co2 and CO2 PC seperately they still produced the same numbers that can be seen under the variable n in the
data above.

**Question 2**

```{r, message=FALSE}
total_CO2<- gapminder %>% 
  group_by(iso3, time) %>% 
  filter(co2 != "NA", name != "NA") %>% 
  summarise(avgCO2 = mean(co2)) %>% 
  filter(iso3 == "USA" | iso3 == "CHN" | iso3 == "IND" | iso3 == "VNM" | iso3 == "KOR")
  
 

total_CO2 %>% 
  ggplot(aes(time, avgCO2, col = factor(iso3))) +
  geom_line() +
  labs(x = "Time (yr)", y = "Average CO2 Emmission (kt)", col = "Countries",
       title = "Average CO2 Emission Over Year")
```



The graph suggests that out of the 5 countries chosen, China on average emits the highest amount
of Co2 after around 2005. China has the greatest positive trend throughout the years with a spike
in emissions after year 2000. The US started with the greatest amount of average co2 emission but 
has not had any major spikes, and is number 2 after 2000. While India, Korea, and Vietnam
emits the least amount of Co2 throughout the years in this group. 



**Question 3**

```{r, message=FALSE}
percapita <- gapminder %>%
  group_by(iso3, time) %>% 
  filter(co2_PC != "NA", name != "NA") %>%
  summarise(avgcapita = mean(co2_PC)) %>% 
  filter(iso3 == "USA" | iso3 == "CHN" | iso3 == "IND" | iso3 == "VNM" | iso3 == "KOR")


percapita %>%
  ggplot(aes(time, avgcapita, col = factor(iso3))) +
  geom_line() +
  labs(x = "Time (yr)", y = "Average CO2 Emmission Per Capita (kt)", col = "Countries", 
       title = "Average CO2 Emission CP Over Year")

```



The graph suggests that the USA has the greatest average Co2 emission per-capita from 1960 to
2019, however there is a steady negative trend with the US releasing less CO2 per-capita in 2019 
than compared to previous years. With the other countries, there is a positive trend where 
they have been increasing CO2 emissions per-capital. 




**Question 4**

```{r, message=FALSE}
gapminder %>%
  group_by(region, time) %>% 
  filter(time == 1960 | time ==2016) %>% 
  filter(co2_PC != "NA", region !="NA") %>%
  summarise(avgCO2_PC = mean(co2_PC))

```

Looking at he the average CO2 emission per capita for each continent, the data shows that in Africa,
Asia, Europe, and Oceania the average Co2 PC increased from 1960 to 2016. However, in the Americas,
the average CO2 PC actually decreased from 2960 to 2016.




**Question 5**

```{r, message=FALSE}
gapminder %>%
  group_by(region, time) %>% 
  filter(time == 1960 | time == 2016) %>% 
  filter(co2_PC != "NA", region !="NA") %>%
  summarise(avgCO2_PC = unique(mean(co2_PC))) %>% 
  
  ggplot() +
  geom_col(aes(x = region, y = avgCO2_PC, fill = factor(time)), position = "dodge") +
  labs(title = "Average CO2 PC emission per Continent", 
           x = "Continent", y = "Average CO2 PC Emission (kt)", fill = "Year")

```


The bar chart suggests that in Africa, Asia, Europe, and Oceania the average CO2 emission per-capita 
were higher in 2016 than compared to 1960. However, in the Americas, average CO2 emission per-capita
was lower than it was in 1960.






**Question 6**

```{r, message=FALSE}
gapminder %>% 
  group_by(region) %>% 
  filter(time == 2016) %>% 
  filter(name != "NA", co2_PC != "NA", region != "NA") %>% 
  select(name, co2_PC) %>% 
  filter(rank(co2_PC) < 4 | rank(desc(co2_PC)) < 4) %>% 
  arrange(region, co2_PC)


```


_The year 2019 does not work for this problem,_
_I had to switch the year to 2016, and had checked with other TAs during their OH_



**Part 4 - GDP Per Capita:**

**Question 1**

```{r, message=FALSE}
gapminder %>% 
  group_by(name) %>% 
  select(region, time, GDP_PC, lifeExpectancy, totalPopulation) %>% 
  filter(time == 1960) %>% 
  filter(GDP_PC != "NA", lifeExpectancy != "NA", name != "NA", totalPopulation != "NA") %>% 
  ggplot() +
  geom_point(aes(GDP_PC, lifeExpectancy, color = region, size = totalPopulation)) +
  labs(title = "1960 GDP PerCap. vs Life Expectancy", x = "GDP Per Capita (Constant 2010 US$)", 
       y = "Life Expectancy (Yr)", color ="Continent", size = "Country \n Size")

```



The scatter plot suggests that as GDP Per Capita increases, so does the life expectancy. 
Additionally, the plot depicts where continents are on the scale. For instance, we can see
that Africa is towards the left end with lower life expectancy and lower GDP CP while Europe
is on the right side of the plot with higher life expectancy and higher GDP CP. It also specifies
how large a country is by the size of the dot. Larger dot indicates a larger country and smaller
dot indicates a smaller country. 





**Question 2**

```{r, message=FALSE}
gapminder %>% 
  group_by(name) %>% 
  select(region, time, GDP_PC, lifeExpectancy, totalPopulation) %>% 
  filter(time == 2019) %>% 
  filter(GDP_PC != "NA", lifeExpectancy != "NA", name != "NA", totalPopulation != "NA") %>% 
  ggplot() +
  geom_point(aes(GDP_PC, lifeExpectancy, color = region, size = totalPopulation)) +
  labs(title = "2019 GDP PerCap. vs Life Expectancy", x = "GDP Per Capita (Constant 2010 US$)", 
       y = "Life Expectancy (Yr)", color ="Continent", size = "Country \n Size")
```




The scatter plot suggests that there are many countries with low GDP PC but high Life Expectancy.
Also, that some countries from Europe, Asia, Oceania, and Americas have the highest GDP PC and highest life
expectancy, although that does not apply to all countries. Many to all of countries in Africa have some of the lowest GDP PC
compared to others. 




**Question 3**


Comparing the two plots, from 1960 to 2019, there has definetly been a shift in the gap between GDP CP and
life expectancy in countries. More countries have greater life expectancy, and there seem to be less of a gap between countries.
For instance, in 1960, Europe was dominatly in the lead, however in 2019, although there is still a gap for some European countries, 
more countries in Asia and Americas can compare to Europe. Lastly, in 2019 there are more dots in the scatter plot
compared to the scatter plot of 1960 indicating that there are more countries being included into the data.





**Question 4**

```{r, message=FALSE}
gapminder %>%
  group_by(region, time) %>% 
  select(lifeExpectancy, time) %>% 
  filter(lifeExpectancy !="NA", region != "NA") %>% 
  filter(time == 1960 | time == 2019) %>% 
  summarise(avg_life_exp_reg = (mean(lifeExpectancy)))

```


Yes, the results fit the figures. In the scatter plots, we can see a general shift in countries having an increase
in life expectancy and looking at the data, each continent from 1960 to 2019 has an increase in their average life expectancy.




**Question 5**

```{r, message=FALSE}
gapminder %>%
  select(region, lifeExpectancy, time) %>% 
  filter(!is.na(lifeExpectancy), !is.na(region)) %>% 
  filter(time == 1960 | time == 2019) %>% 
  group_by(region, time) %>%
  summarise(avg_exp = mean(lifeExpectancy)) %>% 
  summarise(prev = lag(avg_exp),
         growth = avg_exp - prev) %>% 
  filter(!is.na(prev), !is.na(growth)) %>% 
  arrange(desc(growth))
```

What the data is showing us is the growth of the average life expectancy in each continent from 1960 to 2019.
Taking the mean of the average life expectancy in each continent and finding the difference between the two years. 




**Question 6**

```{r, message=FALSE}
gapminder %>% 
  filter(time == 1960 | time == 2019) %>% 
  filter(!is.na(region), !is.na (name), !is.na(GDP_PC)) %>% 
  ggplot() +
  geom_histogram(aes(GDP_PC, fill = factor(time)), position = "dodge") +
  labs(title = "GDP Per Capita for years of 1960 & 2019", x = "GDP_PC (Constant US 2010$)",
       y = "Count", fill = "Year")

```



The histogram above is showing us how many countries had what GDP per capita for the the years of 
1960 (red) and 2019 (blue). We can see that a majority of countries had a GDP PC of under 20,000 in both
1960 and 2019, with few having GCP PC over 30,000. Under 5 countries had GDP PC over 30,000 in 1960 but 
there are around 40 countries with a GDP PC of over 30,000 in 2019. 



**Question 7**

```{r, message=FALSE}
gapminder %>%
  select(name, time, lifeExpectancy) %>% 
  filter(!is.na(time), !is.na(lifeExpectancy), !is.na(name)) %>%
  filter(time == 1960) %>% 
  mutate(ranking = rank(desc(lifeExpectancy))) %>%
  arrange(desc(rank(lifeExpectancy))) %>% 
  filter(name == "United States of America")
```




```{r, message=FALSE}
gapminder %>%
  select(name, time, lifeExpectancy) %>% 
  filter(!is.na(time), !is.na(lifeExpectancy), !is.na(name)) %>%
  filter(time == 2019 ) %>% 
  mutate(ranking = rank(desc(lifeExpectancy))) %>%
  arrange(desc(rank(lifeExpectancy))) %>% 
  filter(name == "United States of America")
```




**Question 8**

```{r, message=FALSE}
gapminder %>% 
  select(name, time, lifeExpectancy) %>% 
  filter(!is.na(time), !is.na(lifeExpectancy), !is.na(name)) %>%
  filter(time == 1960) %>% 
  mutate(ranking = rank(desc(lifeExpectancy))) %>%
  filter(name == "United States of America") %>% 
  mutate(adjusted = ranking/188)
```


```{r, message=FALSE}
gapminder %>% 
  select(name, time, lifeExpectancy) %>% 
  filter(!is.na(time), !is.na(lifeExpectancy), !is.na(name)) %>%
  filter(time == 2019) %>% 
  mutate(ranking = rank(desc(lifeExpectancy))) %>%
  filter(name == "United States of America") %>% 
  mutate(adjusted = (ranking/196))
```




_For this Assignment, I spent a total of 9 hours solving the problems_




















